Raw Earthquake Analysis
========================

------

### Load libraries

```{r}
library(maps)
library(Hmisc)
```



------

## Processing

Download the data, read the data in and save the raw data as an RDA file

```{r loadData,cache=TRUE}
getwd()
download.file("http://earthquake.usgs.gov/earthquakes/catalogs/eqs7day-M1.txt",destfile="../../data/earthquakes.csv",method="curl")
dateDownloaded <- date()
dateDownloaded
quakesRaw <- read.csv('../../data/earthquakes.csv')
save(quakesRaw,dateDownloaded,file="../../data/quakesRaw.rda")
```


### Look at the data set

```{r, dependson="loadData"}
head(quakesRaw)
summary(quakesRaw)
sapply(quakesRaw[1,],class)
```

### Find out about missing values
```{r, dependson="loadData"}
sum(is.na(quakesRaw))
```

### Find minimum and maximum times
```{r, dependson="loadData"}
timeF = strptime(quakesRaw$Datetime,format="%A, %B %e, %Y %H:%M:%S")
min(timeF)
max(timeF)

```


------

## Exploratory analysis

### Make some univariate plots/summaries

```{r, dependson="loadData"}
hist(quakesRaw$Magnitude,breaks=100)
quantile(quakesRaw$Magnitude)
hist(quakesRaw$Depth,breaks=100)
quantile(quakesRaw$Depth)
hist(quakesRaw$Lat,breaks=100)
hist(quakesRaw$Lon,breaks=100)
```

### Make some univariate tables
```{r, dependson="loadData"}
table(quakesRaw$Src)
table(quakesRaw$Version)
table(quakesRaw$Region)
length(unique(quakesRaw$NST))
length(unique(quakesRaw$Eqid))
```


### Plot the earthquakes on the globe
```{r, dependson="loadData"}
map("world")
points(quakesRaw$Lon,quakesRaw$Lat,pch=19,col="blue")
```


### Plot the earthquakes on the globe/sized by relative magnitude 
```{r, dependson="loadData"}
map("world")
points(quakesRaw$Lon,quakesRaw$Lat,pch=19,col="blue",cex=quakesRaw$Magnitude/max(quakesRaw$Magnitude))
```


### Plot the earthquakes on the globe/sized by relative depth
```{r, dependson="loadData"}
map("world")
points(quakesRaw$Lon,quakesRaw$Lat,pch=19,col="blue",cex=quakesRaw$Depth/max(quakesRaw$Depth))
```


### Plot depth versus magnitude
```{r, dependson="loadData"}
plot(quakesRaw$Depth,quakesRaw$Magnitude,pch=19)
```

### Looks weird, let's try a transform - need to add one to avoid log(0)
```{r, dependson="loadData"}
summary(log10(quakesRaw$Depth),useNA="ifany")
summary(log10(quakesRaw$Depth+1))
quakesRaw$log10Depth <- log10(quakesRaw$Depth + 1)
```

### Plot w/transform
```{r, dependson="loadData"}
plot(quakesRaw$log10Depth,quakesRaw$Magnitude,pch=19)
```



### Color by Latitute 
```{r, dependson="loadData"}
latCut = cut2(quakesRaw$Lat,g=5)
plot(quakesRaw$log10Depth,quakesRaw$Magnitude,pch=19,col=latCut)
```

Might be a lot of black in the top right quadrant

### Color by longitude 

```{r, dependson="loadData"}
lonCut = cut2(quakesRaw$Lon,g=5)
plot(quakesRaw$log10Depth,quakesRaw$Magnitude,pch=19,col=lonCut)
```
Definitely a lot of light blue in the upper right hand quadrant


### Color by NST


```{r, dependson="loadData"}
nstCut = cut2(quakesRaw$NST,g=5)
plot(quakesRaw$log10Depth,quakesRaw$Magnitude,pch=19,col=nstCut)
```


### Check out relationship with Src
```{r, dependson="loadData"}
boxplot(quakesRaw$log10Depth ~ quakesRaw$Src)
boxplot(quakesRaw$Magnitude ~ quakesRaw$Src)
```

### How do lat/lon correlated with source

```{r, dependson="loadData"}

boxplot(quakesRaw$Lat ~quakesRaw$Src)
boxplot(quakesRaw$Lon ~quakesRaw$Src)

```



### Get the formatted time, see earthquakes over time

```{r, dependson="loadData"}
timeF = strptime(quakesRaw$Datetime,format="%A, %B %e, %Y %H:%M:%S")
plot(timeF, quakesRaw$log10Depth)
plot(timeF,quakesRaw$Magnitude)
```

------

## Modeling


## Fit a basic model relating depth to time

```{r, dependson="loadData"}
lm1 <- lm(quakesRaw$Depth ~ quakesRaw$Magnitude)
summary(lm1)
```


## Plot the observed ('black') and fitted ('red') points

```{r, dependson="loadData"}
lm1 <- lm(quakesRaw$Magnitude ~ quakesRaw$Depth)
plot(quakesRaw$Depth,quakesRaw$Magnitude,pch=19)
points(quakesRaw$Depth,lm1$fitted,pch=19,col="red")
```


## Look at residuals versus observations, residuals versus fitted values

```{r, dependson="loadData"}
lm1 <- lm(quakesRaw$Magnitude ~ quakesRaw$Depth)
par(mfrow=c(1,2))
plot(quakesRaw$Depth,lm1$residuals,pch=19)
plot(lm1$fitted,lm1$residuals,pch=19)
```

## Try the transform

```{r, dependson="loadData"}
lm2 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth)
plot(quakesRaw$log10Depth,quakesRaw$Magnitude,pch=19)
points(quakesRaw$log10Depth,lm2$fitted,pch=19,col="red")
```


## Plot the residuals versus the observed and fitted (little better here)

```{r, dependson="loadData"}
lm2 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm2$residuals,pch=19)
plot(lm2$fitted,lm2$residuals,pch=19)
```

## What if we color residuals by lat (see a pattern)

```{r, dependson="loadData"}
lm2 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth)
latCut = cut2(quakesRaw$Lat,g=5)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm2$residuals,pch=19,col=latCut)
plot(lm2$fitted,lm2$residuals,pch=19,col=latCut)
```


## What if we color residuals by lon (see a pattern)

```{r, dependson="loadData"}
lm2 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth)
lonCut = cut2(quakesRaw$Lon,g=5)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm2$residuals,pch=19,col=lonCut)
plot(lm2$fitted,lm2$residuals,pch=19,col=lonCut)
```

## Now try fitting a new model with Lat in there

```{r, dependson="loadData"}
latCut = cut2(quakesRaw$Lat,g=5)
lm3 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth + latCut)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm3$residuals,pch=19,col=latCut)
plot(lm3$fitted,lm3$residuals,pch=19,col=latCut)
```

## How about Lat/Lon

```{r, dependson="loadData"}
latCut = cut2(quakesRaw$Lat,g=5)
lonCut = cut2(quakesRaw$Lon,g=5)
lm4 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth + latCut + lonCut)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm4$residuals,pch=19,col=latCut)
plot(lm4$fitted,lm4$residuals,pch=19,col=latCut)
```


## Color that model by NST

```{r, dependson="loadData"}
latCut = cut2(quakesRaw$Lat,g=5)
lonCut = cut2(quakesRaw$Lon,g=5)
nstCut = cut2(quakesRaw$NST,g=5)
lm4 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth + latCut + lonCut)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm4$residuals,pch=19,col=nstCut)
plot(lm4$fitted,lm4$residuals,pch=19,col=nstCut)
```

## Include NST 

```{r lm5Chunk, dependson="loadData"}
latCut = cut2(quakesRaw$Lat,g=5)
lonCut = cut2(quakesRaw$Lon,g=5)
nstCut = cut2(quakesRaw$NST,g=5)
lm5 <- lm(quakesRaw$Magnitude ~ quakesRaw$log10Depth + latCut + lonCut + nstCut)
par(mfrow=c(1,2))
plot(quakesRaw$log10Depth,lm5$residuals,pch=19,col=nstCut)
plot(lm5$fitted,lm5$residuals,pch=19,col=nstCut)
```



## Let's use model 5

```{r, dependson="lm5Chunk"}
summary(lm5)
confint(lm5)
```

## Let's plot fitted versus observed magnitudes on the map

```{r, dependson="lm5Chunk"}
par(mfrow=c(2,1))
map("world")
lm5fitted <- lm5$fitted
points(quakesRaw$Lon,quakesRaw$Lat,pch=19,col="blue",cex=lm5fitted/max(lm5fitted))
map("world")
points(quakesRaw$Lon,quakesRaw$Lat,pch=19,col="blue",cex=quakesRaw$Magnitude/max(quakesRaw$Magnitude))
```


