Final Earthquake Analysis
========================

Note: This analysis was created on a Mac. Some functions may be altered to be able to run on Windows machines. 

------

## Preliminaries

### Load libraries/data/create new variables

```{r loadData}

# Load libraries
library(maps)
library(Hmisc)
library(RColorBrewer)

# Load data
load("../../data/quakesRaw.rda")

# Define relevant variables - making numeric variables into factors when they should be.
quakesRaw$latCut <- cut2(quakesRaw$Lat,g=5)
quakesRaw$lonCut <- cut2(quakesRaw$Lon,g=5)
quakesRaw$nstCut <- cut2(quakesRaw$NST,g=5)
quakesRaw$log10Depth <- log10(quakesRaw$Depth + 1)
quakesRaw$time <- strptime(quakesRaw$Datetime,format="%A, %B %e, %Y %H:%M:%S")

## This is the data set we will use
quakes <- quakesRaw
```


------

## Exploratory analysis

### Get minimum and maximum times and date downloaded (Methods/Data Collection)

```{r, dependson="loadData"}
min(quakes$time)
max(quakes$time)
dateDownloaded
```

### Find number of missing values/check ranges (Results paragraph 1)

```{r, dependson="loadData"}
sum(is.na(quakes))
summary(quakes)
```
Latitude, longitude are within normal ranges. Magnitude has nothing above 7, depth is within the defined range. 


### Look at patterns over time (Results paragraph 1)
```{r, dependson="loadData"}
plot(quakes$time,quakes$Magnitude,pch=19)
plot(quakes$time,quakes$Depth,pch=19)
```
There does not appear to be a time trend in either variable. 


### Look at distribution of magnitudes (Results paragraph 2)
```{r, dependson="loadData"}
mean(quakes$Magnitude < 3)
mean(quakes$Magnitude > 3 & quakes$Magnitude < 5)
```
Most earthquakes are small (< 3) or medium (>3 and < 5)

### Look at distribution of depths (Results paragraph 2)

```{r, dependson="loadData"}
hist(quakes$Depth,col="grey")
hist(quakes$log10Depth,col="grey")
```

-------

## Modeling 

### Fit a model with no adjustment (results - paragraph 3)

```{r, dependson="loadData"}
# Fit model with no adjustment variable
lmNoAdjust <- lm(quakes$Magnitude ~ quakes$log10Depth)

# Plot residuals, colored by different variables (latitude, longitude, number of sites observing the quake)
par(mfrow=c(1,3))
plot(quakes$log10Depth,lmNoAdjust$residuals,col=quakes$latCut,pch=19)
plot(quakes$log10Depth,lmNoAdjust$residuals,col=quakes$lonCut,pch=19)
plot(quakes$log10Depth,lmNoAdjust$residuals,col=quakes$nstCut,pch=19)
```
It appears there are some non-random patterns here.


### Now fit a model with factor adjustment for latitude, longitude, and number of sites (results - paragraph 3)

```{r lmFinalChunk, dependson="loadData"}
lmFinal <- lm(quakes$Magnitude ~ quakes$log10Depth + quakes$latCut + quakes$lonCut + quakes$NST)
par(mfrow=c(1,3))
plot(quakes$log10Depth,lmFinal$residuals,col=quakes$latCut,pch=19)
plot(quakes$log10Depth,lmFinal$residuals,col=quakes$lonCut,pch=19)
plot(quakes$log10Depth,lmFinal$residuals,col=quakes$nstCut,pch=19)

```
Still some clumpiness of color, but much better than it was. 

## Get the estimates and confidence intervals

```{r, dependson="lmFinalChunk"}
## The estimate from summary
summary(lmFinal)

## The confidence interval from confint
confint(lmFinal)
```

-------

## Figure making
```{r, dependson="lmFinalChunk"}
lmNoAdjust <- lm(quakes$Magnitude ~ quakes$log10Depth)

## Set up a function that makes colors prettier
mypar <- function(a=1,b=1,brewer.n=8,brewer.name="Dark2",...){
 par(mar=c(2.5,2.5,1.6,1.1),mgp=c(1.5,.5,0))
 par(mfrow=c(a,b),...)
 palette(brewer.pal(brewer.n,brewer.name))
}

## Set size of axes
cx = 1.3

## Save figure to pdf file
pdf(file="../../figures/finalfigure.pdf", height=4, width=3*4)
mypar(mfrow=c(1,3))

hist(quakes$Depth,breaks=100,col=1,xlab="Depth (km)",ylab="Frequency",main="",cex.axis=cx,cex.lab=cx)
plot(quakes$log10Depth,lmNoAdjust$residuals,col=quakes$latCut,pch=19,xlab="Log Base 10 Depth (km)", ylab="No Adjustment Residuals",cex.axis=cx,cex.lab=cx)
plot(quakes$log10Depth,lmFinal$residuals,col=quakes$latCut,pch=19,xlab="Log Base 10 Depth (km)", ylab="Full Model Residuals",cex.axis=cx,cex.lab=cx)
```
